---
title: "Data cleaning"
author: "Maria Braeuner"
date: "11/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r LoadPackages, results="hide", warning=FALSE, message=FALSE}

library(tidyverse) 
library(lubridate) #date functions

```

The first part of this document details the steps to clean, transform & prepare the raw data from Cyclistic for analysis. The second part is the initial exploratory analysis. Final analysis/results in CS1_Cyclistic_Report.Rmd. 

### Preparing
#### Import Data

```{r LoadData, results="hide", warning=FALSE, message=FALSE}

#Load Data

Nov20 <- read_csv("Data/202011-divvy-tripdata.csv")
Dec20 <- read_csv("Data/202012-divvy-tripdata.csv")
Jan21 <- read_csv("Data/202101-divvy-tripdata.csv")
Feb21 <- read_csv("Data/202102-divvy-tripdata.csv")
Mar21 <- read_csv("Data/202103-divvy-tripdata.csv")
Apr21 <- read_csv("Data/202104-divvy-tripdata.csv")
May21 <- read_csv("Data/202105-divvy-tripdata.csv")
Jun21 <- read_csv("Data/202106-divvy-tripdata.csv")
Jul21 <- read_csv("Data/202107-divvy-tripdata.csv")
Ago21 <- read_csv("Data/202108-divvy-tripdata.csv")
Sep21 <- read_csv("Data/202109-divvy-tripdata.csv")
Oct21 <- read_csv("Data/202110-divvy-tripdata.csv")

#Check all df's have the same structure: check if all column names
#are the same (e.g. would make it easier to merge into one dataframe)
colnames(Nov20) == colnames(Dec20)
colnames(Nov20) == colnames(Jan21)
colnames(Nov20) == colnames(Feb21)
colnames(Nov20) == colnames(Mar21)
colnames(Nov20) == colnames(Apr21)
colnames(Nov20) == colnames(May21)
colnames(Nov20) == colnames(Jun21)
colnames(Nov20) == colnames(Jul21)
colnames(Nov20) == colnames(Ago21)
colnames(Nov20) == colnames(Sep21)
colnames(Nov20) == colnames(Oct21)
# all TRUE

```

The data available for each df: `r colnames(Nov20)`


### Merge Data 
```{r MergeData, results="hide", warning=FALSE, message=FALSE}

#Merge into one df
cyclistic12 <- rbind(Nov20, Dec20, Jan21, Feb21,
                     Mar21, Apr21, May21, Jun21,
                     Jul21, Ago21, Sep21, Oct21)
#Check:
head(cyclistic12)
tail(cyclistic12)
dim(cyclistic12) # 5,378,834 x 13

```

```{r tib1} 

head(cyclistic12)

```

### Clean & prepare data for analysis

```{r CleanData, message=FALSE, warning=FALSE, results="hide"}

#check for data types
glimpse(cyclistic12) #rideable_type & member_casual are as chr, change to factors
cyclistic12$rideable_type <- as.factor(cyclistic12$rideable_type)
cyclistic12$member_casual <- as.factor(cyclistic12$member_casual)
glimpse(cyclistic12) #ok

summary(cyclistic12)
#ended_at includes 3 days of November 2021 (remove)
#end_lat & end_lang have 4831 NA's
#rideable_type categories: classic_bike, docked_bike, electric_bike
#member_casual categories: member & casual

length(unique(cyclistic12$start_station_name)) #check how many stations (815)
length(unique(cyclistic12$end_station_name)) #check how many stations (812)

#filter out the extra days of Nov2021
cyclistic <- cyclistic12 %>%
  filter(ended_at <= "2021-11-01 00:00:00")

summary(cyclistic)
dim(cyclistic) #5,378.531 x 13 - ok

#Add new columns for: Day, Month, Year, DayOfWeek (dow), time of day (tod) & tod_char, Season, ride_length
cyclistic$date <- as.Date(cyclistic$started_at) #yyyy-mm-dd
cyclistic$month <- format(as.Date(cyclistic$date), "%m")
cyclistic$month <- as.numeric(cyclistic$month) #helps to create Seasons column;
cyclistic$day <- format(as.Date(cyclistic$date), "%d")
cyclistic$year <- format(as.Date(cyclistic$date), "%Y")
cyclistic$dow <- format(as.Date(cyclistic$date), "%A")
cyclistic$tod <- format(cyclistic$started_at, "%H:%M:%S")
cyclistic$ride_length <- difftime(cyclistic$ended_at, cyclistic$started_at) #in seconds
cyclistic$ride_length <- as.numeric(cyclistic$ride_length)
cyclistic$ride_length_min <- cyclistic$ride_length/60 #in min

cyclistic$season <- "Winter"
cyclistic$season[cyclistic$month>2&cyclistic$month<6] <- "Spring"
cyclistic$season[cyclistic$month>5&cyclistic$month<9] <- "Summer"
cyclistic$season[cyclistic$month>8&cyclistic$month<12] <- "Autumn"
cyclistic$season <- as.factor(cyclistic$season)
summary(cyclistic$season)

# Reference used here: 
##Morning=[05:00-11:59] ; Afternoon=[12:00-17:59] ; Evening=[18:00-21:59] ; Night=[22:00-04:59]

tod_char <- format(cyclistic$started_at, "%H")
tod_char <- as.numeric(tod_char)
cyclistic$tod_char <- "Night"
cyclistic$tod_char[tod_char<12&tod_char>=5] <- "Morning"
cyclistic$tod_char[tod_char>=18&tod_char<22] <- "Evening"
cyclistic$tod_char[tod_char>=12&tod_char<18] <- "Afternoon"
cyclistic$tod_char <- as.factor(cyclistic$tod_char)

summary(cyclistic$tod_char)
glimpse(cyclistic)

summary(cyclistic) #there are negative ride_length vals.

cyclistic[cyclistic$ride_length_min <0, ] # 1393 negatives; started_at & ended_at could be inverted, but not sure, better ignore.

##Exclude these rows.
Cyclistic_Data <- cyclistic[!cyclistic$ride_length <0, ]

dim(cyclistic) # 5378531 x 23
dim(Cyclistic_Data) # 5377138 x 23
# 5378531 - 5377138 = 1393 = ok

#ride_id should not have duplicates:
length(Cyclistic_Data$ride_id) #5377138
n_distinct(Cyclistic_Data$ride_id) #5377138 ok!

max(Cyclistic_Data$ride_length_min) #55944.14, this is over 38 days.
#how many ride lenghts exceed 24 hours?
Cyclistic_Data %>%
  summarize(weird_lengths = which(ride_length_min>1440))
#there are 3,800 trips of over 24 hours
# who's doing 24+? Check:
long_rides <- Cyclistic_Data %>%
  group_by(member_casual) %>%
  summarize(Long_Ride_Length = which(ride_length_min>1440))

long_rides <- long_rides %>%
  group_by(member_casual) %>%
  summarize(n = n(),
            mean_duration_minutes = mean(Long_Ride_Length))

#Station names have some that were tests and some NAs
##Remove testing rows. Leave NAs for now (some don't have station name but lat,lon)
CyclisticData <- Cyclistic_Data[!grepl("TEST",Cyclistic_Data$start_station_name), ]

CyclisticData <- CyclisticData[!grepl("TEST",CyclisticData$end_station_name), ]

CyclisticData <- CyclisticData[!grepl("TEST",CyclisticData$start_station_id), ]

CyclisticData <- CyclisticData[!grepl("TEST",CyclisticData$end_station_id), ]

rideable_type_check <- CyclisticData %>%
  group_by(month, year) %>%
  select(rideable_type, month, year) %>%
  count(rideable_type)

rideable_type_check

#order dow by dow instead of alphabetically:
CyclisticData$dow <- ordered(CyclisticData$dow, levels=c("Monday",
                                                         "Tuesday",
                                                         "Wednesday",
                                                         "Thursday",
                                                         "Friday",
                                                         "Saturday",
                                                         "Sunday"))


#order season
CyclisticData$season <- ordered(CyclisticData$season,
                                levels=c("Winter",
                                         "Spring",
                                         "Summer",
                                         "Autumn"))


```

Keep in mind for analysis:  


* Rideable_type for November 2020 includes only 2 categories: docked and electric; "classic_bike" appears from December 2020 onward.
* Reference used for "Time of Day" (tod_char): 
  + Morning = [05:00-11:59]
  + Afternoon = [12:00-17:59]
  + Evening = [18:00-21:59]
  + Night = [22:00-04:59]
* There are 3,800 trips of over 24 hours still included in the dataset (mean_duration_minutes = mean() of the length of all trips over 1440 minutes (24 hours), in minutes):
```{r long_rides }

long_rides

```

Save cleaned data into a new .csv file.
```{r SaveData, include=FALSE}

#CyclisticData <- write.csv(CyclisticData, "Data/CyclisticData.csv")

```


# Exploratory Analysis

```{r, tidy=TRUE}

head(CyclisticData)

```

How do casual customers and members differ in # of rides & average ride duration (in minutes)?
```{r EA1, warning=FALSE, message=FALSE, tidy=TRUE}

## Ride length ($ride_length &/or $ride_length_min)

CyclisticData %>%
  group_by(member_casual) %>%
  summarise(no_rides = n(),
           avg_duration_min = mean(ride_length_min))

```

How do casual customers and members differ in # of rides & average ride duration (in minutes) by ride_type used?
```{r EA2, warning=FALSE, message=FALSE, tidy=TRUE}
## Type of ride they use ($rideable_type)
CyclisticData %>%
  group_by(member_casual, rideable_type) %>%
  summarise(no_rides = n(),
           avg_duration_min = mean(ride_length_min))

```

How do casual customers and members differ in # of rides & average ride duration (in minutes) by season?

```{r EA3, warning=FALSE, message=FALSE, tidy=TRUE}
## Use by $season
CyclisticData %>%
  group_by(member_casual, season) %>%
  summarise(no_rides = n(),
           avg_duration_min = mean(ride_length_min))

```

How do casual customers and members differ in # of rides & average ride duration (in minutes) by time of day?
```{r EA4, warning=FALSE, message=FALSE, tidy=TRUE}
## Use by $tod ($tod_char)
CyclisticData %>%
  group_by(member_casual, tod_char) %>%
  summarise(no_rides = n(),
           avg_duration_min = mean(ride_length_min))

```

How do casual customers and members differ in # of rides & average ride duration (in minutes) by day of the week?

```{r EA5, warning=FALSE, message=FALSE, tidy=TRUE}
## Use by $dow
CyclisticData %>%
  group_by(member_casual, dow) %>%
  summarise(no_rides = n(),
           avg_duration_min = mean(ride_length_min))

```


```{r EAviz, warning=FALSE, message=FALSE, tidy=TRUE}

CyclisticData %>%
  group_by(member_casual, dow) %>%
  summarise(no_rides = n(),
           avg_duration_min = mean(ride_length_min)) %>%
  ggplot(mapping = aes(x = dow, y = avg_duration_min, fill = member_casual)) +
  geom_col(position="dodge")


CyclisticData %>%
  group_by(member_casual, dow) %>%
  summarise(no_rides = n(),
           avg_duration_min = mean(ride_length_min)) %>%
  ggplot(mapping = aes(x = dow, y = no_rides, fill = member_casual)) +
  geom_col(position="dodge")


CyclisticData %>%
  group_by(member_casual, tod_char) %>%
  summarise(no_rides = n(),
           avg_duration_min = mean(ride_length_min)) %>%
  ggplot(mapping = aes(x = tod_char, y = avg_duration_min, fill = member_casual)) +
  geom_col(position="dodge")

CyclisticData %>%
  group_by(member_casual, tod_char) %>%
  summarise(no_rides = n(),
           avg_duration_min = mean(ride_length_min)) %>%
  ggplot(mapping = aes(x = tod_char, y = no_rides, fill = member_casual)) +
  geom_col(position="dodge")



CyclisticData %>%
  group_by(member_casual, season) %>%
  summarise(no_rides = n(),
           avg_duration_min = mean(ride_length_min)) %>%
  ggplot(mapping = aes(x = season, y = no_rides, fill = member_casual)) +
  geom_col(position="dodge")




```

